---
title: "Using AnVIL VRS Toolkit in R"
author: 
- name: Marcel Ramos
  affiliation: >
    CUNY Graduate School of Public Health and Health Policy, New York, NY USA
output: BiocStyle::html_document
date: "`r format(Sys.time(), '%B %d, %Y')`"
vignette: >
  %\VignetteIndexEntry{AnVILVRS Introduction}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
package: AnVILVRS
---

# AnVILVRS

# Introduction

The AnVILVRS package provides an R interface to the
[AnVIL VRS Toolkit](https://pypi.org/project/vrs-anvil-toolkit/), a Python
library for working with the [Global Alliance for Genomics and Health (GA4GH)
Variation Representation Specification (VRS)](https://vrs.ga4gh.org/en/stable/)
standard. The package allows users to translate variant identifiers from various
formats (e.g., gnomAD, SPDI, HGVS, Beacon) into GA4GH VRS Allele IDs and vice
versa. Additionally, it provides functionality to retrieve allele frequency data
from the 1000 Genomes Project based on VRS Allele IDs.

# Installation

To use the AnVILVRS package, you need to have Python installed on your system.
The package requires Python 3.11, so ensure that you have this version
installed.

Install the AnVILVRS package from Bioconductor with:

```{r install, eval=FALSE}
if (!requireNamespace("BiocManager", quietly = TRUE))
    install.packages("BiocManager")

BiocManager::install("AnVILVRS")
```

# Loading and Setup

Load the AnVILVRS package and the reticulate package for Python integration:

```{r load}
library(AnVILVRS)
library(reticulate)
```

After installing the package, you need to set up a Python virtual environment
with the required dependencies. You can do this by running the following command
in R:

```{r setup, eval=FALSE}
install_AnVILVRS(envname = "vrs_env")
```

# Usage

Once the virtual environment is set up, you can use the AnVILVRS package to
translate variant identifiers and retrieve allele frequency data.
First, load the package and activate the virtual environment:

```{r}
use_virtualenv("vrs_env", required = TRUE)
```

## Translating Variant Identifiers

You can translate variant identifiers from various formats into GA4GH VRS Allele
IDs using the `get_vrs_id` function. Supported formats include "gnomad",
"spdi", "hgvs", and "beacon".

```{r translate}
get_vrs_id("chr7-87509329-A-G", "gnomad")
get_vrs_id("NC_000005.10:80656509:C:TT", "spdi")
get_vrs_id("NC_000005.10:g.80656510delinsTT", "hgvs")
get_vrs_id("5 : 80656489 C > T", "beacon")
```

## Allele Object Retrieval

You can also retrieve the VRS Allele object using the `get_vrs_allele` function:

```{r allele}
allele <- get_vrs_allele("5 : 80656489 C > T", "beacon")
allele
```

## Variant Retrieval from Allele Object

You can convert a VRS Allele object back to a variant identifier in a specified
format using the `get_variant_from_allele` function:

```{r variant_from_allele}
get_variant_from_allele(allele, "hgvs")
```

## Retrieving Allele Frequency Data

To retrieve allele frequency data from the 1000 Genomes Project based on a VRS
Allele ID, use the `get_caf` function. You will need a bgz zipped VCF file
containing 1000 Genomes Project variants with allele frequency annotations and
its corresponding index file.

