---
title: "Using AnVIL VRS Toolkit in R"
author: 
- name: Marcel Ramos
  affiliation: >
    CUNY Graduate School of Public Health and Health Policy, New York, NY USA
output: BiocStyle::html_document
date: "`r format(Sys.time(), '%B %d, %Y')`"
vignette: >
  %\VignetteIndexEntry{AnVILVRS Introduction}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
package: AnVILVRS
---

# AnVILVRS

# Introduction

The AnVILVRS package provides an R interface to the
[AnVIL VRS Toolkit](https://pypi.org/project/vrs-anvil-toolkit/), a Python
library for working with the [Global Alliance for Genomics and Health (GA4GH)
Variation Representation Specification (VRS)](https://vrs.ga4gh.org/en/stable/)
standard. The package allows users to translate variant identifiers from various
formats (e.g., gnomAD, SPDI, HGVS, Beacon) into GA4GH VRS Allele IDs and vice
versa. Additionally, it provides functionality to retrieve allele frequency data
from the 1000 Genomes Project based on VRS Allele IDs.

# Installation

To use the AnVILVRS package, you need to have Python installed on your system.
The package requires Python 3.11, so ensure that you have this version
installed.

Install the AnVILVRS package from Bioconductor with:

```{r install, eval=FALSE}
if (!requireNamespace("BiocManager", quietly = TRUE))
    install.packages("BiocManager")

BiocManager::install("AnVILVRS")
```

# Loading and Setup

Load the AnVILVRS package and the reticulate package for Python integration:

```{r load}
library(AnVILVRS)
library(reticulate)
```

After installing the package, you need to set up a Python virtual environment
with the required dependencies. You can do this by running the following command
in R:

```{r setup, eval=FALSE}
install_AnVILVRS(envname = "vrs_env")
```

# Usage

Once the virtual environment is set up, you can use the AnVILVRS package to
translate variant identifiers and retrieve allele frequency data.
First, load the package and activate the virtual environment:

```{r}
use_virtualenv("vrs_env", required = TRUE)
```

## Translating Variant Identifiers

You can translate variant identifiers from various formats into GA4GH VRS Allele
IDs using the `get_vrs_id` function. Supported formats include "gnomad",
"spdi", "hgvs", and "beacon".

```{r translate}
get_vrs_id("chr7-87509329-A-G", "gnomad")
get_vrs_id("NC_000005.10:80656509:C:TT", "spdi")
get_vrs_id("NC_000005.10:g.80656510delinsTT", "hgvs")
get_vrs_id("5 : 80656489 C > T", "beacon")
```

## Allele Object Retrieval

You can also retrieve the VRS Allele object using the `get_vrs_allele` function:

```{r allele}
allele <- get_vrs_allele("5 : 80656489 C > T", "beacon")
allele
```

## Variant Retrieval from Allele Object

You can convert a VRS Allele object back to a variant identifier in a specified
format using the `get_variant_from_allele` function:

```{r variant_from_allele}
get_variant_from_allele(allele, "hgvs")
```

## Retrieving Allele Frequency Data

To retrieve allele frequency data from the 1000 Genomes Project based on a VRS
Allele ID, use the `get_caf` function. You will need a `.gz` zipped VCF file
containing 1000 Genomes Project variants with allele frequency annotations and
its corresponding index file.

```{r, eval=FALSE}
use_virtualenv("vrs_env", required = TRUE)
vcf <- "../vrs_anvil_toolkit/tests/fixtures/1kGP.chr1.1000.vrs.vcf.gz"
vcf_index <- "../1000g_chr1_index.db"
variant_id <- "chr1-20094-TAA-T"
vrs_id <- get_vrs_id(variant_id, "gnomad")
pop_desc <- get_pop_descriptor()
get_caf(
  vrs_id, vcf, vcf_index, "USA",
  pop_desc_file = pop_desc, toolkit_dir = "../vrs_anvil_toolkit"
)
```

## Data download functions

The `get_pop_descriptor` function downloads the population descriptor file
from a known Google Storage URI to the BiocFileCache. This file is used in the
`get_caf` function to map population codes.

```{r download}
library(readr)
pop_desc <- get_pop_descriptor()
read_tsv(
    file = pop_desc,
    col_types = cols(
        population_descriptor_id = col_character(),
        population_descriptor = col_character(),
        subject_id = col_character(),
        country_of_recruitment = col_character(),
        population_label = col_character()
    )
)
```

The `get_seqrepo` function downloads a SeqRepo archive from a specified URI to
a local directory. This is useful for setting up the SeqRepo database required
by the AnVIL VRS Toolkit.

```{r seqrepo, eval=FALSE}
get_seqrepo(destdir = tempdir())
```
